### 2014-11-13 coordination.tex
Several corrections. Reduce size of texttt.

The SSSP program is concise and declarative but its performance depends on the order in which nodes are
executed. If nodes with greater distances are prioritized over other nodes, the
program will generate more \texttt{relax} facts since it will take longer to
reach the shortest distances. From Fig.~\ref{fig:shortest_path_program}, it is
clear that the best scheduling is the following: \texttt{@1},
\texttt{@3}, \texttt{@2} and then \texttt{@4}, where only 4 \texttt{relax}
facts are generated. If we had decided to process nodes in order
\texttt{@1}, \texttt{@2}, \texttt{@4}, \texttt{@3}, \texttt{@4},
\texttt{@2}, then 6 \texttt{relax} facts would have been generated.
The optimal solution for SSSP is to schedule the node with the
shortest distance, which is essentially the Dijkstra shortest path
algorithm~\cite{Dijkstra}. Note how it is possible to change the nature of
the algorithm by simply changing the order of node computation, but still
retain the declarative nature of the program.

@@

\newcommand{\code}[1]{\texttt{\small{#1}}}

\begin{tightitemize}
   \item \code{set-priority(node A, float F)}: This sets the
   temporary priority of \texttt{A} to \texttt{F}. If \texttt{A} has a priority
   \texttt{F'}, we only change the priority if \texttt{F} is higher. The programmer
   can decide if priorities are to be ordered in ascending or descending order.
   \item \code{add-priority(node A, float F)}: Increases,
   temporarily, the priority of node \texttt{A} by \texttt{F}.
   \item \code{remove-priority(node A)}: Removes the temporary priority from node
   \texttt{A}.
   \item \code{schedule-next(node A)}: Changes the temporary priority of node
   \texttt{A} to be 1 unit higher than the current highest priority.
   \item \code{set-default-priority(node, float)}: Sets the default
   priority of the node.
   \item \code{stop-program(node A)}: Immediately stops the
   execution of the whole program.
\end{tightitemize}

CLM provides the sensing fact \code{priority(node A, node B, float P)} in order to sense the
priority \texttt{P} of node \texttt{B} (at node \texttt{A}).
Sensing facts are only used in the body of the rules in order
to fetch information from the runtime system.
Note that when sensing facts are used to prove new facts, they are re-derived automatically.
~\footnote{The observant reader will notice that all the coordination facts are
linear and when created in the body of a rule will be consumed. The system
creates the necessary code to re-derive them without programmer interaction.
Likewise, \texttt{set-priority} and \texttt{set-default-priority} update the
value of \texttt{priority} facts by consuming and re-deriving them but this is
done automatically by the runtime system.}

@@

\begin{tightitemize}
   \item \code{set-cpu(node A, int T)}: Moves node \texttt{A} to thread
   \texttt{T}.
   \item \code{set-affinity(node A, node B)}: Places node \texttt{B} in
   the thread of node \texttt{A}.
   \item \code{set-moving(node A)}: Allows node \texttt{A} to move freely
   between threads.
   \item \code{set-static(node A)}: Forces node \texttt{A} to stay in the
   same thread indefinitely.
\end{tightitemize}

@@

\begin{tightitemize}
   \item \code{cpu-id(node A, node B, int T)}: Store at node \texttt{A} which
   thread \texttt{T}, \texttt{B} is actually running on.
   \item \code{moving(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is allowed
   to move between threads.
   \item \code{static(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is not
   allowed to move between threads.
\end{tightitemize}

@@

\begin{tightdescription}
   \item[\texttt{priority @order ORDER.}] \texttt{ORDER} can be either \texttt{asc} or \texttt{desc}. This defines if node's are to be selected by the smallest or the greatest priority, respectively.
   \item[\texttt{priority @initial P.}] The \texttt{initial} statement informs the runtime system that all nodes must start with priority $P$. Alternatively, the programmer can define an \texttt{set-priority(A, P)} axiom.
   \item[\texttt{priority @static.}] The \texttt{static} priority tells the runtime system that the partition of nodes among workers is to be used until the end of program. 
\end{tightdescription}

>>>>

The SSSP program is concise and declarative but its performance depends on the order in which nodes are
executed. If nodes with greater distances are prioritized over other nodes, the
program will generate more \mytt{relax} facts since it will take longer to
reach the shortest distances. From Fig.~\ref{fig:shortest_path_program}, it is
clear that the best scheduling is the following: \mytt{@1},
\mytt{@3}, \mytt{@2} and then \mytt{@4}, where only 4 \mytt{relax}
facts are generated. If we had decided to process nodes in order
\mytt{@1}, \mytt{@2}, \mytt{@4}, \mytt{@3}, \mytt{@4},
\mytt{@2}, then 6 \mytt{relax} facts would have been generated.
The optimal solution for SSSP is to schedule the node with the
shortest distance, which is essentially the Dijkstra shortest path
algorithm~\cite{Dijkstra}. Note how it is possible to change the nature of
the algorithm by simply changing the order of node computation, but still
retain the declarative nature of the program.

@@

\newcommand{\code}[1]{\mytt{\small{#1}}}

\begin{tightitemize}
   \item \code{set-priority(node A, float F)}: This sets the
   temporary priority of \mytt{A} to \mytt{F}. If \mytt{A} has a priority
   \mytt{F'}, we only change the priority if \mytt{F} is higher. The programmer
   can decide if priorities are to be ordered in ascending or descending order.
   \item \code{add-priority(node A, float F)}: Increases,
   temporarily, the priority of node \mytt{A} by \mytt{F}.
   \item \code{remove-priority(node A)}: Removes the temporary priority from node
   \mytt{A}.
   \item \code{schedule-next(node A)}: Changes the temporary priority of node
   \mytt{A} to be 1 unit higher than the current highest priority.
   \item \code{set-default-priority(node, float)}: Sets the default
   priority of the node.
   \item \code{stop-program(node A)}: Immediately stops the
   execution of the whole program.
\end{tightitemize}

CLM provides the sensing fact \code{priority(node A, node B, float P)} in order to sense the
priority \mytt{P} of node \mytt{B} (at node \mytt{A}).
Sensing facts are only used in the body of the rules in order
to fetch information from the runtime system.
Note that when sensing facts are used to prove new facts, they are re-derived automatically.
~\footnote{The observant reader will notice that all the coordination facts are
linear and when created in the body of a rule will be consumed. The system
creates the necessary code to re-derive them without programmer interaction.
Likewise, \mytt{set-priority} and \mytt{set-default-priority} update the
value of \mytt{priority} facts by consuming and re-deriving them but this is
done automatically by the runtime system.}

@@

\begin{tightitemize}
   \item \code{set-cpu(node A, int T)}: Moves node \mytt{A} to thread
   \mytt{T}.
   \item \code{set-affinity(node A, node B)}: Places node \mytt{B} in
   the thread of node \mytt{A}.
   \item \code{set-moving(node A)}: Allows node \mytt{A} to move freely
   between threads.
   \item \code{set-static(node A)}: Forces node \mytt{A} to stay in the
   same thread indefinitely.
\end{tightitemize}

@@

\begin{tightitemize}
   \item \code{cpu-id(node A, node B, int T)}: Store at node \mytt{A} which
   thread \mytt{T}, \mytt{B} is actually running on.
   \item \code{moving(node A, node B)}: Fact available at node \mytt{A} if \mytt{B} is allowed
   to move between threads.
   \item \code{static(node A, node B)}: Fact available at node \mytt{A} if \mytt{B} is not
   allowed to move between threads.
\end{tightitemize}

@@

\begin{tightdescription}
   \item[\mytt{priority @order ORDER.}] \mytt{ORDER} can be either \mytt{asc} or \mytt{desc}. This defines if node's are to be selected by the smallest or the greatest priority, respectively.
   \item[\mytt{priority @initial P.}] The \mytt{initial} statement informs the runtime system that all nodes must start with priority $P$. Alternatively, the programmer can define an \mytt{set-priority(A, P)} axiom.
   \item[\mytt{priority @static.}] The \mytt{static} priority tells the runtime system that the partition of nodes among workers is to be used until the end of program. 
\end{tightdescription}

### 2014-11-13 heat-transfer.tex
Several corrections. Reduce size of texttt.

Fig.~\ref{code:ht} shows the HT rules that send new heat values to
neighbor nodes. In the first rule we added \texttt{add-priority} to increase the priority of the neighbor
nodes if the current node has a large $\delta$. The idea is to prioritize the
computation of heat values of nodes (using \texttt{update}) that have a neighbor
that changed significantly. Multiple \texttt{add-priority} facts will
increase the priority of a node so that nodes with multiple large deltas will
have more priority.

@@

\begin{topfig}
   \begin{center}
      \includegraphics[width=5.5cm]{results/new-heat-transfer-80}
   \end{center}
   \scap{results:ht}{Experimental results for the HT program. The coordinated version
      is, on average, 30\% faster than the regular version although it has a
      slightly worse scalability due to a reduction in work available.}
\end{topfig}

>>>>

Fig.~\ref{code:ht} shows the HT rules that send new heat values to
neighbor nodes. In the first rule we added \mytt{add-priority} to increase the priority of the neighbor
nodes if the current node has a large $\delta$. The idea is to prioritize the
computation of heat values of nodes (using \mytt{update}) that have a neighbor
that changed significantly. Multiple \mytt{add-priority} facts will
increase the priority of a node so that nodes with multiple large deltas will
have more priority.

@@

\begin{topfig}
   \begin{center}
      \includegraphics[width=\figsize]{results/new-heat-transfer-80}
   \end{center}
   \scap{results:ht}{Experimental results for the HT program. The coordinated version
      is, on average, 30\% faster than the regular version although it has a
      slightly worse scalability due to a reduction in work available.}
\end{topfig}

### 2014-11-13 implementation.tex
Several corrections. Reduce size of texttt.

\paragraph{Coordination directives}
Coordination directives are compiled in two different ways, depending on whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the directive \texttt{node-priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of being added to the database as facts.
Semantically, action facts are like any other. However, since they are
immediately used by the machine, there is no need to store them in the
database, therefore avoiding unnecessary allocations and deallocations.

@@

The \texttt{next} and \texttt{prev} pointers of the regular queue are
part of the node structure in order to save space. These pointers are
also used as the index in the priority queue and current priority,
respectively.

>>>>

\paragraph{Coordination directives}
Coordination directives are compiled in two different ways, depending on whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the fact \mytt{priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of being added to the database as facts.
Semantically, action facts are like any other. However, since they are
immediately used by the machine, there is no need to store them in the
database, therefore avoiding unnecessary allocations and deallocations.

@@

The \mytt{next} and \mytt{prev} pointers of the regular queue are
part of the node structure in order to save space. These pointers are
also used as the index in the priority queue and current priority,
respectively.

### 2014-11-13 language.tex
Several corrections. Reduce size of texttt.

Coordinated Linear Meld (CLM) is a forward-chaining linear logic programming
language based on LM~\cite{cruz-iclp14}.
Programs consist of a set of \emph{rules} and a \emph{database of facts}.
Rules have the form \texttt{a(X), b(Y) -o c(X, Y)} and can be read as follows: if fact \texttt{a(X)}
and fact \texttt{b(Y)} exist in the database then \texttt{c(X, Y)} is added to
the database. The expression \texttt{a(X), b(Y)} is called the \emph{body} of
the rule and \texttt{c(X, Y)} is the \emph{head} of the rule.
A fact is a predicate, e.g., \texttt{a}, \texttt{b} or \texttt{c}, and its
associated tuple of values, e.g., the concrete values of \texttt{X} and
\texttt{Y}. Since CLM uses linear logic as its foundation, we distinguish between
\emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
process of deriving a rule, while persistent facts are not.
Program execution starts by adding the \emph{axioms} (the initial facts of the
program) to the database. Next, rules are recursively applied and the
database is updated by adding new facts or deleting facts used during rule derivation.
When no more rules are applicable, the program terminates.

CLM has been designed for writing programs that operate on graphs. To achieve
concurrency, CLM partitions the database by using the first argument of each fact. The first argument has
type \emph{node} and represents a node in the graph. For example,
the fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while
fact \texttt{p(@2)} is stored in node \texttt{@2}. CLM restricts the body of
every rule to facts with the same node so that nodes can derive rules
independently. The head of the rule may refer
to any node as long as that node is referred to somewhere in the body. This allows
\scare{communication} between nodes during rule derivation, since a node may
\scare{send} a fact to another node.  Rule restrictions in turn make CLM
implicitly parallel because nodes are able to compute independently. This makes
CLM non-deterministic since nodes can be picked to run in any order, affecting
which rules are applied and which facts are deleted or derived.

@@

The SSSP program starts (lines 1-3) with the declaration of the predicates.
Predicates specify the facts used in the program. The first predicate,
\texttt{edge}, is a persistent predicate that describes the
relationship between the nodes of the graph, where the third argument
represents the weight of the edge (the \texttt{route} modifier
informs the compiler that the \texttt{edge} predicate
determines the structure of the graph).  The predicates
\texttt{shortest} and \texttt{relax} are specified as linear facts
and thus are deleted when deriving new facts.  The algorithm computes
the shortest distance from node \texttt{@1} to all other nodes in the
graph. Every node has a \texttt{shortest} fact that is improved with
new \texttt{relax} facts.  Lines 5-9 declare the axioms of the
program: \texttt{edge} facts describe the graph; \texttt{shortest(A,
+00, [])} is the initial shortest distance (infinity) for all
nodes; and \texttt{relax(@1, 0, [@1])} starts the algorithm by
setting the initial distance from \texttt{@1} to \texttt{@1} to be 0.

@@

\begin{dblfig}
\vspace*{-2ex}
\begin{center}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest2}}
  \hspace{0.4cm}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest3}}
  \hspace{0.4cm}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest8}}
\end{center}
\scap{fig:shortest_path_program}{Graphical representation of the SSSP program. (a) represents the
   program after propagating initial distance at node \texttt{@1}, followed by
   (b) where the first rule is applied in node \texttt{@3}. (c)
   represents the state of the final program, where all the shortest paths
   have been computed.}
\vspace*{-1ex}
\end{dblfig}

The first rule of the program (lines 11-14) reads as following: if the current
\texttt{shortest} path \texttt{P1} with distance \texttt{D1} is larger
than a new path \texttt{relax} with distance \texttt{D2}, then replace the
current shortest path with \texttt{D2}, delete the new \texttt{relax} path and
propagate new paths to the neighbors (lines 13-14) using a \emph{comprehension}.
The comprehension iterates over the edges of node \texttt{A} and derives a new
\texttt{relax} fact for each node \texttt{B} with the distance \texttt{D2 + W},
where \texttt{W} is the weight of the edge. For
example, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
\texttt{@1} where two new \texttt{relax} facts are derived at node \texttt{@2}
and \texttt{@3}. Fig.~\ref{fig:shortest_path_program}~(b) is the result after
applying the same rule but at node \texttt{2}.

The second rule of the program (lines 16-17) is read as following: if the
current shortest path \texttt{D1} is shorter than the new path \texttt{D2} then
delete the new \texttt{relax} fact and keep the current shortest path.

There are many opportunities for concurrency in the SSSP program. For instance,
after applying rule 1 in Fig.~\ref{fig:shortest_path_program}~(a), it is
possible to either apply rules in either node \texttt{@2} or node
\texttt{@3}. This decision depends largely on implementation factors such
as node partitioning and number of threads in the system.  Still, it is
easy to prove that no matter the scheduling used, the final result, as
presented in Fig.~\ref{fig:shortest_path_program}~(c), is achieved.

>>>>

Coordinated Linear Meld (CLM) is a forward-chaining linear logic programming
language based on LM~\cite{cruz-iclp14}.
Programs consist of a set of \emph{rules} and a \emph{database of facts}.
Rules have the form \mytt{a(X), b(Y) -o c(X, Y)} and can be read as follows: if fact \mytt{a(X)}
and fact \mytt{b(Y)} exist in the database then \mytt{c(X, Y)} is added to
the database. The expression \mytt{a(X), b(Y)} is called the \emph{body} of
the rule and \mytt{c(X, Y)} is the \emph{head} of the rule.
A fact is a predicate, e.g., \mytt{a}, \mytt{b} or \mytt{c}, and its
associated tuple of values, e.g., the concrete values of \mytt{X} and
\mytt{Y}. Since CLM uses linear logic as its foundation, we distinguish between
\emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
process of deriving a rule, while persistent facts are not.
Program execution starts by adding the \emph{axioms} (the initial facts of the
program) to the database. Next, rules are recursively applied and the
database is updated by adding new facts or deleting facts used during rule derivation.
When no more rules are applicable, the program terminates.

CLM has been designed for writing programs that operate on graphs. To achieve
concurrency, CLM partitions the database by using the first argument of each fact. The first argument has
type \emph{node} and represents a node in the graph. For example,
the fact \mytt{f(@1, 2)} is stored in node \mytt{@1}, while
fact \mytt{p(@2)} is stored in node \mytt{@2}. CLM restricts the body of
every rule to facts with the same node so that nodes can derive rules
independently. The head of the rule may refer
to any node as long as that node is referred to somewhere in the body. This allows
\scare{communication} between nodes during rule derivation, since a node may
\scare{send} a fact to another node.  Rule restrictions in turn make CLM
implicitly parallel because nodes are able to compute independently. This makes
CLM non-deterministic since nodes can be picked to run in any order, affecting
which rules are applied and which facts are deleted or derived.

@@

The SSSP program starts (lines 1-3) with the declaration of the predicates.
Predicates specify the facts used in the program. The first predicate,
\mytt{edge}, is a persistent predicate that describes the
relationship between the nodes of the graph, where the third argument
represents the weight of the edge (the \mytt{route} modifier
informs the compiler that the \mytt{edge} predicate
determines the structure of the graph).  The predicates
\mytt{shortest} and \mytt{relax} are specified as linear facts
and thus are deleted when deriving new facts.  The algorithm computes
the shortest distance from node \mytt{@1} to all other nodes in the
graph. Every node has a \mytt{shortest} fact that is improved with
new \mytt{relax} facts.  Lines 5-9 declare the axioms of the
program: \mytt{edge} facts describe the graph; \mytt{shortest(A,
+00, [])} is the initial shortest distance (infinity) for all
nodes; and \mytt{relax(@1, 0, [@1])} starts the algorithm by
setting the initial distance from \mytt{@1} to \mytt{@1} to be 0.

@@

\begin{dblfig}
\vspace*{-2ex}
\begin{center}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest2}}
  \hspace{0.4cm}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest3}}
  \hspace{0.4cm}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest8}}
\end{center}
\scap{fig:shortest_path_program}{Graphical representation of the SSSP program. (a) represents the
   program after propagating initial distance at node \mytt{@1}, followed by
   (b) where the first rule is applied in node \mytt{@2}. (c)
   represents the state of the final program, where all the shortest paths
   have been computed.}
\vspace*{-1ex}
\end{dblfig}

The first rule of the program (lines 11-14) reads as following: if the current
\mytt{shortest} path \mytt{P1} with distance \mytt{D1} is larger
than a new path \mytt{relax} with distance \mytt{D2}, then replace the
current shortest path with \mytt{D2}, delete the new \mytt{relax} path and
propagate new paths to the neighbors (lines 13-14) using a \emph{comprehension}.
The comprehension iterates over the edges of node \mytt{A} and derives a new
\mytt{relax} fact for each node \mytt{B} with the distance \mytt{D2 + W},
where \mytt{W} is the weight of the edge. For
example, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
\mytt{@1} where two new \mytt{relax} facts are derived at node \mytt{@2}
and \mytt{@3}. Fig.~\ref{fig:shortest_path_program}~(b) is the result after
applying the same rule but at node \mytt{2}.

The second rule of the program (lines 16-17) is read as following: if the
current shortest path \mytt{D1} is shorter than the new path \mytt{D2} then
delete the new \mytt{relax} fact and keep the current shortest path.

There are many opportunities for concurrency in the SSSP program. For instance,
after applying rule 1 in Fig.~\ref{fig:shortest_path_program}~(a), it is
possible to either apply rules in either node \mytt{@2} or node
\mytt{@3}. This decision depends largely on implementation factors such
as node partitioning and number of threads in the system.  Still, it is
easy to prove that no matter the scheduling used, the final result, as
presented in Fig.~\ref{fig:shortest_path_program}~(c), is achieved.

### 2014-11-13 locks.tex
Several corrections. Reduce size of texttt.

\begin{dbltab}
   \begin{center}
\footnotesize\begin{tabular}{ | l || c | c | c | c | c | c | c | c | c | c |}
\hline
\multirow{2}{*}{\textbf{Program}} & \multicolumn{5}{c |}{\textbf{Queue
   Instructions}} & \multicolumn{5}{c |}{\textbf{Facts Derived}} \\ \cline{2-11}
& 1 & 2 & 4 & 8 & 16 & 1 & 2 & 4 & 8 & 16\\ \hline
\hline
SSSP - Regular & 199.1K & 203.1K & 202.7K & 203.8K & 206.4K & 3.3M & 3.0M & 3.4M & 3.5M & 3.3M \\
SSSP - Coordinated & 959.0K & 924.3K & 891.0K & 966.3K & 1.0M & 2.0M & 2.3M & 2.3M & 2.2M & 2.3M \\
SSSP - Buffered & 1.0M & 1.0M & 1.1M & 1.0M & 1.1M & 2.0M & 2.1M & 2.3M & 2.3M & 2.4M \\
\hline
HT - Regular & 25.6M & 33.3M & 34.8M & 36.1M & 34.8M & 235.7M & 298.3M & 312.1M & 317.5M & 310.3M \\
HT - Coordinated & 51.6M & 55.2M & 85.9M & 79.9M & 75.6M & 129.2M & 183.4M & 254.1M & 242.0M & 252.4M \\
HT - Local-Only & 51.6M & 54.9M & 81.0M & 76.7M & 67.3M & 129.2M & 180.8M & 221.5M & 217.6M & 207.4M \\
\hline
LBP - Regular & 15.7M & 17.8M & 16.0M & 11.3M & 8.5M & 160.2M & 171.0M & 149.9M & 107.5M & 85.1M \\
SBP - Coordinated & 97.2M & 97.4M & 90.0M & 84.3M & 79.2M & 76.8M & 80.7M & 79.2M & 79.0M & 79.3M \\
\hline \hline
MiniMax - Regular & 16.4M & 18.3M & 20.5M & 19.4M & 20.9M & 76.4M & 76.4M & 76.4M & 76.4M & 76.4M \\
MiniMax - Coordinated & 52.0M & 52.0M & 52.0M & 52.0M & 52.0M & 76.4M & 76.4M & 76.4M & 76.4M & 76.4M \\
\hline
N Queens - Regular & 8.4K & 14.4K & 17.7K & 16.6K & 19.3K & 167.1M & 167.1M & 167.1M & 167.1M & 167.1M \\
N Queens - Coordinated & 57.1M & 24.7M & 1.1M & 141.1K & 41.6K & 167.1M & 167.1M & 167.1M & 167.1M & 167.1M \\
N Queens - Static & 57.1M & 47.2M & 3.4M & 1.5M & 209.2K & 167.1M & 167.1M & 167.1M & 167.1M & 167.1M \\
\hline
\end{tabular}
\vspace*{.5ex}
\end{center}
\scap{queue_instructions}{Total number of queue \scare{instructions} and facts derived per program. An
   operation to a regular queue counts as 1 instruction, while an operation to
   a priority queue costs the number of
   \emph{percolate-up}/\emph{percolate-down} steps performed.}
\end{dbltab}

@@

As can be seen in Table~\ref{queue_instructions} all the coordinated
versions of the programs require significantly more queue operations.
The data for this table comes from recording the number of \emph{queue instructions} and the number of facts derived.
Queue instructions represent the
number of normal queue operations executed (each costs 1) plus the
number of \emph{percolate-up}/\emph{percolate-down} operations
executed for each manipulation of the priority queue.  As expected,
programs which rely on prioritizing the order of execution to obtain
better asymptotic complexity (e.g., SSSP) improve in overall running
time at the cost of more queue operations.  In all cases, coordination
adds a significant number of total queue operations, but the resulting
overhead is more than compensated for by an improved schedule and reduction
in number of facts produced. For MiniMax and N Queens, the number of facts derived
is the same for all configurations, as expected. The performance seen for those programs
arise from reduced memory usage and improved memory locality.

@@

\begin{topfig}
   \begin{center}
      \includegraphics[width=6.5cm]{results/locks/sssp-locks.png}
   \end{center}
  \scap{locks:SSSP}{Locking statistics for SSSP.
     \texttt{Coordinated} refers to the program in
     Fig.~\ref{results:sssp_uspowergrid}, while \texttt{Buffered} is a more
     optimized version of the same program where coordination operations are
     buffered before application.}
\end{topfig}

@@

We next investigate the effect of coordination on the synchronization
overhead that comes from the use of more locks.  We measured the
\emph{basic locks}, the number of locks acquired
in the original LM virtual machine and \emph{coordination locks}, the
number of locks acquired to perform coordination operations.  We also
measured how often \texttt{lock()} operations that fail to acquired
the lock on the first attempt.  Lock failures happen so infrequently
(less than 1\% of the time) that they cannot be seen on the figures.

@@

Fig.~\ref{locks:HT} shows the locking behavior of the heat transfer
program.  The behavior is representative of the other programs. In general, there
is a drop in the total number of locks acquired and coordination locks
represent a small fraction of the total number of locks acquired.  The \textbf{Local-Only}
version further reduces locking by keeping more derivations local
to a thread.

@@

\iffalse

  The
SSSP program is The first program we measured was SSSP, represented in
Fig.~\ref{results:sssp_uspowergrid}. In addition to the Regular and
Coordinated versions, we have the Buffered version where we optimized
the virtual machine to buffer coordination actions before applying
them. The idea is to avoid applying multiple \texttt{set-priority}
operations to the same node during a single node execution. We thus
gather the higher priority values and then perform coordination after
the node has executed.  Although the number of locks is significantly
reduced, we did not see a similar performance improvement, since the
Buffered version is only 5\% faster than the Coordinated version.

In Fig.~\ref{locks:HT} we show the lock statistics for HT. We measured the
regular version, the coordinated version and the local-only version as in
Fig.~\ref{results:ht}. There is a slight reduction in number of locks with
coordination and then a further reduction in the local-only version because less
facts are exchanged between threads.

Finally, in Fig.~\ref{locks:LBP} we present lock statistics for LBP and SBP. LBP
is represented as the Regular bar, while SBP is represented as the Coordinated
bar. We notice that LBP has quite a lot of locking for 1 thread since LBP
is an asynchronous algorithm and tends to be a slow sequential algorithm,
requiring more derivations when using 1 thread. The numbers go down as more
threads are used. For SBP, there is a consistent amount of locking without
little variations between the number of threads used. This is because threads
interact very little with each other and the amount of work to do is the same
for all threads.

\begin{topfig}
   \begin{center}
      \includegraphics[width=6.5cm]{results/locks/bp.png}
   \end{center}
  \scap{locks:LBP}{Locking statistics for LBP and SBP. Locking is constant for
     SBP because threads do not interact with each other.}
\end{topfig}

\paragraph{Queue Instructions}

Table~\ref{queue_instructions} presents the total queue instructions for several
programs. All the coordinated programs require far more queue instructions than
the regular version. For instance, the coordinated SSSP program requires 4 to 5
times more instructions to perform queue operations. Interestingly, the buffered
version requires even more instructions although it performs less overall
coordination. It is likely that this is due to increased fact computation and
subsequent queue manipulation because coordination operations are buffered.
We argue that the 5\% performance improvement in the
buffered version is due mostly to reduced locking.

For the HT program, the coordinated version requires twice more queue
instructions than the regular version. As expected, the local-only version
requires less queue operations because the number of derived facts is reduced.
The MiniMax program also sees twice as many queue instructions using
coordination. This does not happen with the N Queens program, where the number
of queue instructions is far greater when using coordination, especially when
using a small number of threads. We argue that this happens because higher
priority nodes at the bottom are often added to the priority queue, resulting in
many \emph{percolate-up} operations.

These experiments indicate that handling coordination has a small cost in terms
of synchronization but some cost in terms of queue operations with
up to 5 times more queue instructions when using priority queues.
Since coordination reduces the number of facts derived due to better scheduling
decisions, the costs of manipulating priority queues become irrelevant.
\fi

>>>>

\begin{dbltab}
   \begin{center}
\footnotesize\begin{tabular}{ | l || c | c | c | c | c | c | c | c | c | c |}
\hline
\multirow{2}{*}{\textbf{Program}} & \multicolumn{5}{c |}{\textbf{Queue
   Operations}} & \multicolumn{5}{c |}{\textbf{Facts Derived}} \\ \cline{2-11}
& 1 & 2 & 4 & 8 & 16 & 1 & 2 & 4 & 8 & 16\\ \hline
\hline
SSSP - Regular & 199.1K & 203.1K & 202.7K & 203.8K & 206.4K & 3.3M & 3.0M & 3.4M & 3.5M & 3.3M \\
SSSP - Coordinated & 959.0K & 924.3K & 891.0K & 966.3K & 1.0M & 2.0M & 2.3M & 2.3M & 2.2M & 2.3M \\
SSSP - Buffered & 1.0M & 1.0M & 1.1M & 1.0M & 1.1M & 2.0M & 2.1M & 2.3M & 2.3M & 2.4M \\
\hline
HT - Regular & 25.6M & 33.3M & 34.8M & 36.1M & 34.8M & 235.7M & 298.3M & 312.1M & 317.5M & 310.3M \\
HT - Coordinated & 51.6M & 55.2M & 85.9M & 79.9M & 75.6M & 129.2M & 183.4M & 254.1M & 242.0M & 252.4M \\
HT - Local-Only & 51.6M & 54.9M & 81.0M & 76.7M & 67.3M & 129.2M & 180.8M & 221.5M & 217.6M & 207.4M \\
\hline
LBP - Regular & 15.7M & 17.8M & 16.0M & 11.3M & 8.5M & 160.2M & 171.0M & 149.9M & 107.5M & 85.1M \\
SBP - Coordinated & 97.2M & 97.4M & 90.0M & 84.3M & 79.2M & 76.8M & 80.7M & 79.2M & 79.0M & 79.3M \\
\hline \hline
MiniMax - Regular & 16.4M & 18.3M & 20.5M & 19.4M & 20.9M & 76.4M & 76.4M & 76.4M & 76.4M & 76.4M \\
MiniMax - Coordinated & 52.0M & 52.0M & 52.0M & 52.0M & 52.0M & 76.4M & 76.4M & 76.4M & 76.4M & 76.4M \\
\hline
N Queens - Regular & 8.4K & 14.4K & 17.7K & 16.6K & 19.3K & 167.1M & 167.1M & 167.1M & 167.1M & 167.1M \\
N Queens - Coordinated & 57.1M & 24.7M & 1.1M & 141.1K & 41.6K & 167.1M & 167.1M & 167.1M & 167.1M & 167.1M \\
N Queens - Static & 57.1M & 47.2M & 3.4M & 1.5M & 209.2K & 167.1M & 167.1M & 167.1M & 167.1M & 167.1M \\
\hline
\end{tabular}
\vspace*{.5ex}
\end{center}
\scap{queue_instructions}{Total number of queue operations and facts derived per program. An
   operation to a regular queue counts as 1 instruction, while an operation to
   a priority queue costs the number of
   \emph{percolate-up}/\emph{percolate-down} steps performed.}
\end{dbltab}

@@

As can be seen in Table~\ref{queue_instructions} all the coordinated
versions of the programs require significantly more queue operations.
The data for this table comes from recording the number of \emph{queue
operations} and facts derived.
Queue instructions represent the
number of normal queue operations executed (each costs 1) plus the
number of \emph{percolate-up}/\emph{percolate-down} operations
executed for each manipulation of the priority queue.  As expected,
programs which rely on prioritizing the order of execution to obtain
better asymptotic complexity (e.g., SSSP) improve in overall running
time at the cost of more queue operations.  In all cases, coordination
adds a significant number of total queue operations, but the resulting
overhead is more than compensated for by an improved schedule and reduction
in number of facts produced. For MiniMax and N Queens, the number of facts derived
is the same for all configurations, as expected. The performance seen for those programs
arise from reduced memory usage and improved memory locality.

@@

\begin{topfig}
   \begin{center}
      \includegraphics[width=6.5cm]{results/locks/sssp-locks.png}
   \end{center}
  \scap{locks:SSSP}{Locking statistics for SSSP.
     \mytt{Coordinated} refers to the program in
     Fig.~\ref{results:sssp_uspowergrid}, while \mytt{Buffered} is a more
     optimized version of the same program where coordination operations are
     buffered before application.}
\end{topfig}

@@

We next investigate the effect of coordination on the synchronization
overhead that comes from the use of more locks.  We measured the
\emph{basic locks}, the number of locks acquired
in the original LM virtual machine and \emph{coordination locks}, the
number of locks acquired to perform coordination operations.  We also
measured how often \mytt{lock()} operations that fail to acquired
the lock on the first attempt.  Lock failures happen so infrequently
(less than 1\% of the time) that they cannot be seen on the figures.

@@

Fig.~\ref{locks:HT} shows the locking behavior of the heat transfer
program.  The behavior is representative of the other programs. In general, there
is a drop in the total number of locks acquired and coordination locks
represent a small fraction of the total number of locks acquired. \textbf{Local-Only}
further reduces locking by keeping more derivations local to a thread.

@@

### 2014-11-13 minimax.tex
Several corrections. Reduce size of texttt.

The code in Fig.~\ref{minimax:check-end} deals with the tree expansion process.
The first three rules (lines 1-10) deal
with the case where no children nodes are created and the last three rules
(12-29) deal with the cases for that create new nodes. In particular, the two
rules in lines 12-25 generate new nodes using the
\texttt{exists} language construct, which creates a child node 
\texttt{B}. We link \texttt{B} with its parent (\texttt{parent(B, A)})
and kick start the expansion of that node \texttt{B} by adding a \texttt{play}
fact.

@@

We also take advantage of memory locality by using \texttt{set-static} (line
14), so that nodes after a certain level are not stolen by other threads. While
this is not critical for performance in shared memory systems where node
stealing is fairly efficient, we expect that such coordination to be critical in
distributed systems.

>>>>

The code in Fig.~\ref{minimax:check-end} deals with the tree expansion process.
The first three rules (lines 1-10) deal
with the case where no children nodes are created and the last three rules
(12-29) deal with the cases for that create new nodes. In particular, the two
rules in lines 12-25 generate new nodes using the
\mytt{exists} language construct, which creates a child node 
\mytt{B}. We link \mytt{B} with its parent (\mytt{parent(B, A)})
and kick start the expansion of that node \mytt{B} by adding a \mytt{play}
fact.

@@

We also take advantage of memory locality by using \mytt{set-static} (line
14), so that nodes after a certain level are not stolen by other threads. While
this is not critical for performance in shared memory systems where node
stealing is fairly efficient, we expect that such coordination to be critical in
distributed systems.
